---
title: "Potential of Fecal Microbiota for Early-Stage Detection of Colorectal Cancer"
author: "Sally Karimi Kinyua"
output:
  slidy_presentation: default
  ioslides_presentation:
    widescreen: true
    smaller: true
subtitle: A Differential Abundance Analysis of the ZellerG_2014 Cohort
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")

library(curatedMetagenomicData)
library(dplyr)
library(scater)
library(mia)
library(ANCOMBC)
library(lefser)

 ##Installing packages to aid with ANCOMBC
remotes::install_version("CVXR", version = "1.0-15", repos = "https://cran.r-project.org")
packageVersion("CVXR")
library(CVXR)
"solve" %in% ls("package:CVXR")
```

## Background

Colorectal cancer (CRC) develops gradually, typically progressing through
a precursor lesion — the **adenoma** — before becoming invasive cancer.
Several bacterial taxa have been proposed as biomarkers of this
progression, but their consistency across disease stages is still an
active area of research.

This analysis re-examines gut microbiome composition across three stages
of that progression, using shotgun metagenomic data from
**Zeller *et al.*, 2014**, *"Potential of fecal microbiota for early-stage
detection of colorectal cancer,"* published in *Molecular Systems Biology*.


## Research question

**Does gut microbial community composition differ between healthy
controls, adenoma patients, and CRC patients — and if so, which specific
taxa drive that difference?**

Three groups are compared throughout:

| Group     | Description                          | n   |
|-----------|---------------------------------------|-----|
| `control` | Healthy, tumour-free individuals      | 61  |
| `adenoma` | Colorectal adenoma (precursor lesion) | 42  |
| `CRC`     | Colorectal cancer                     | 53  |

`control` is used as the **reference group** throughout, so every result
below is framed as "adenoma vs. healthy" or "CRC vs. healthy."


## Analytical approach

Two complementary differential abundance methods are used:

- **ANCOM-BC2** — a compositionally-aware method that models bias
  directly and works on absolute count-scale data. Generally more
  statistically powered.
- **LEfSe** (Linear discriminant analysis Effect Size) — combines
  non-parametric testing (Kruskal-Wallis, Wilcoxon) with linear
  discriminant analysis to rank taxa by effect size. More conservative,
  widely used for its interpretable effect-size ranking.

Agreement between the two methods is treated as stronger evidence for a
given taxon's association with disease status.

# Data Preparation

## Identifying the dataset

The `curatedMetagenomicData` package hosts hundreds of public microbiome
studies with harmonised metadata, making it possible to reproduce and
extend published analyses like Zeller et al. (2014) directly.

```{r explore-studies}
crc_related_terms <- c("CRC", "adenoma", "carcinoma_surgery_history", "control")

crc_overview <- sampleMetadata |>
  filter(study_condition %in% crc_related_terms) |>
  count(study_name, study_condition)

crc_overview
```

`ZellerG_2014` was selected for having all three disease stages
represented with reasonably balanced sample sizes.
`carcinoma_surgery_history` (present only in a different study) was
excluded throughout, as it reflects a distinct post-surgical population
rather than a comparable disease stage.


## Confirming the sample subset

```{r single-study-meta}
single_study_name <- "ZellerG_2014"

single_meta <- sampleMetadata |>
  filter(study_name == single_study_name,
         study_condition %in% c("control", "adenoma", "CRC"))

table(single_meta$study_condition)
```


## Downloading abundance data

`curatedMetagenomicData()` can return data on two different scales,
controlled by the `counts` argument. Both are needed here, since the two
downstream methods expect different input scales:

- **`counts = FALSE`** → true relative abundance (proportions, summing to
  ~100 per sample) — used for **quality control, diversity, and LEfSe**
- **`counts = TRUE`** → estimated read counts (scaled by sequencing
  depth) — used for **ANCOM-BC2**, which models count-based bias directly

```{r single-study-download-ra}
single_data_ra <- curatedMetagenomicData(
  "2021-03-31.ZellerG_2014.relative_abundance",
  dryrun = FALSE,
  counts = FALSE
)

single_tse_ra <- single_data_ra[[1]]
single_tse_ra <- single_tse_ra[, single_tse_ra$study_condition %in% c("control", "adenoma", "CRC")]

single_tse_ra
```


## Verifying the abundance scale

A quick sanity check confirms the data is on the expected proportional
scale before any analysis begins.

```{r single-study-check-ra}
colSums(assay(single_tse_ra, "relative_abundance"))[1:5]
range(assay(single_tse_ra, "relative_abundance"))
```

Column sums close to 100 confirm true relative abundance, not raw counts.

# Quality Control & Community Structure

## Sequencing depth consistency

Before testing for biological differences, it's important to confirm
there's no systematic technical bias between groups — for example, one
group being sequenced to a different depth than another.

```{r single-qc}
colData(single_tse_ra)$total_reads <- colSums(assay(single_tse_ra, "relative_abundance"))

plotColData(
  single_tse_ra,
  y = "total_reads",
  x = "study_condition",
  colour_by = "study_condition"
)
```

All three groups cluster tightly around 100%, confirming no systematic
QC bias between them.


## Alpha diversity: within-sample richness

Shannon diversity captures how evenly and richly the microbial community
is represented within each individual sample.

```{r single-alpha}
single_tse_ra <- addAlpha(
  single_tse_ra,
  assay.type = "relative_abundance",
  index = "shannon"
)

plotColData(
  single_tse_ra,
  y = "shannon",
  x = "study_condition",
  colour_by = "study_condition"
)
```

Diversity is broadly similar across all three groups — consistent with
the CRC literature, where disease-associated microbiome shifts are
usually driven by specific taxa rather than an overall loss of diversity.


## Beta diversity: between-sample community structure

Bray-Curtis dissimilarity, visualised via Principal Coordinates Analysis
(PCoA), shows whether samples cluster by disease status based on overall
community composition.

```{r single-beta}
single_tse_ra <- runMDS(
  single_tse_ra,
  FUN = vegan::vegdist,
  method = "bray",
  assay.type = "relative_abundance",
  name = "PCoA_Bray"
)

plotReducedDim(
  single_tse_ra,
  dimred = "PCoA_Bray",
  color_by = "study_condition"
)
```

Substantial overlap between groups — again pointing toward taxon-specific
signals rather than a wholesale community shift, which is exactly what
the differential abundance methods below are designed to detect.

# Differential Abundance: ANCOM-BC2

## Preparing the count-scale data

```{r single-study-download-counts}
single_data <- curatedMetagenomicData(
  "2021-03-31.ZellerG_2014.relative_abundance",
  dryrun = FALSE,
  counts = TRUE
)

single_tse <- single_data[[1]]
single_tse <- single_tse[, single_tse$study_condition %in% c("control", "adenoma", "CRC")]

# Set 'control' as the reference level so every comparison reads
# naturally as "vs. healthy control"
single_tse$study_condition <- factor(
  single_tse$study_condition,
  levels = c("control", "adenoma", "CRC")
)

levels(single_tse$study_condition)
```


## Running ANCOM-BC2

```{r single-ancombc2}
single_out <- ancombc2(
  data = single_tse,
  assay_name = "relative_abundance",
  fix_formula = "study_condition"
)

colnames(single_out$res)
```

With three groups, the model produces two sets of coefficients —
`study_conditionadenoma` and `study_conditionCRC` — each contrasted
against the `control` reference.


## Results: adenoma vs. control

```{r single-ancombc2-adenoma}
res_adenoma_vs_control <- single_out$res %>%
  select(
    taxon,
    lfc = lfc_study_conditionadenoma,
    se = se_study_conditionadenoma,
    p_val = p_study_conditionadenoma,
    q_val = q_study_conditionadenoma,
    diff = diff_study_conditionadenoma
  ) %>%
  arrange(q_val)

sum(res_adenoma_vs_control$diff, na.rm = TRUE)  # count of significant taxa
head(res_adenoma_vs_control, 10)
```

**21 taxa** are significantly differentially abundant between adenoma
patients and healthy controls.


## Results: CRC vs. control

```{r single-ancombc2-crc}
res_crc_vs_control <- single_out$res %>%
  select(
    taxon,
    lfc = lfc_study_conditionCRC,
    se = se_study_conditionCRC,
    p_val = p_study_conditionCRC,
    q_val = q_study_conditionCRC,
    diff = diff_study_conditionCRC
  ) %>%
  arrange(q_val)

sum(res_crc_vs_control$diff, na.rm = TRUE)  # count of significant taxa
head(res_crc_vs_control, 10)
```

**38 taxa** are significantly differentially abundant between CRC
patients and healthy controls — nearly double the adenoma comparison,
consistent with more extensive microbial disruption at the more advanced
disease stage.

# Differential Abundance: LEfSe

## A scaling note

`lefser()` internally expects abundances scaled to sum to **1,000,000**
per sample (parts-per-million), a convention inherited from the original
LEfSe tool. Its own `lefser:::relativeAb()` function applies this
scaling correctly — general-purpose relative-abundance transforms (e.g.
from `mia`, which scale to 0–1) are not equivalent here and will cause
the LDA effect-size step to silently return no results.

```{r lefse-adenoma}
single_ra_adenoma <- lefser:::relativeAb(
  single_tse[, single_tse$study_condition %in% c("control", "adenoma")]
)
single_ra_adenoma$study_condition <- droplevels(
  as.factor(single_ra_adenoma$study_condition)
)

res_lefse_adenoma <- lefser(
  single_ra_adenoma,
  classCol = "study_condition",
  lda.threshold = 3,
  trim.names = TRUE
)
lefserPlot(res_lefse_adenoma)
```


## LEfSe: CRC vs. control

```{r lefse-crc}
single_ra_crc <- lefser:::relativeAb(
  single_tse[, single_tse$study_condition %in% c("control", "CRC")]
)
single_ra_crc$study_condition <- droplevels(
  as.factor(single_ra_crc$study_condition)
)

res_lefse_crc <- lefser(
  single_ra_crc,
  classCol = "study_condition",
  lda.threshold = 3,
  trim.names = TRUE
)
lefserPlot(res_lefse_crc)
```

Several taxa enriched in CRC here — including *Fusobacterium* and
*Escherichia coli* — echo well-established findings in the CRC
microbiome literature, and corroborate the ANCOM-BC2 results above
using an independent statistical approach.

# Summary & Interpretation

## Key findings

- **Community-level structure** (alpha and beta diversity) does not
  differ dramatically across control, adenoma, and CRC — the microbiome
  is not globally disrupted at either disease stage.
- **Taxon-level differences are substantial**: ANCOM-BC2 identified
  **21 significant taxa** in adenoma vs. control and **38** in
  CRC vs. control, showing a clear escalation with disease progression.
- **LEfSe independently corroborates** several of the same taxa,
  including CRC-associated enrichment of *Fusobacterium* and
  *Escherichia coli* — consistent with prior literature on CRC-microbiome
  associations.
- Together, these results support the idea that **specific bacterial
  taxa, rather than overall community diversity, are the more sensitive
  signal** for CRC and its precursor lesions.

## Question
How can we perform a multi-group LEfSe analysis to identify differentially abundant features across the three conditions simultaneously?


## Next steps

- Extend this analysis to additional CRC cohorts
  (e.g. `YachidaS_2019`, `ThomasAM_2018a`, `FengQ_2015`) to test whether
  these taxa replicate across independent populations
- Adjust for study-level batch effects when pooling multiple cohorts
- Cross-reference significant taxa against curated microbiome signature
  databases (e.g. BugSigDB) for independent validation


## Thank you

**Sally Karimi Kinyua**

`r Sys.Date()`
