---
title: "DELIVERY MODE ANALYSIS — WampachL_2018"
subtitle: "Vaginal vs. C-section, species-level differential abundance"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

<!--
##########################################################
## DELIVERY MODE ANALYSIS — WampachL_2018
## Vaginal vs. C-section, species-level differential abundance
## Methods: ANCOM-BC2, LEfSe, MAaSLin2 (+ optional corncob, DESeq2)
##########################################################
-->

> **Methods:** ANCOM-BC2, LEfSe, MAaSLin2 (+ optional Corncob, DESeq2)

# Setup

```{r setup}
library(curatedMetagenomicData)
library(dplyr)
library(ANCOMBC)
library(lefser)
library(Maaslin2)
```

# Download Data

```{r step1-download-data}
wampach_data <- curatedMetagenomicData(
  "WampachL_2018.relative_abundance",
  dryrun = FALSE,
  counts = TRUE   # counts scale needed for ANCOM-BC2/MAaSLin2/corncob/DESeq2
)

tse_delivery_sp <- wampach_data[[1]]
```

# Restrict to Samples with Known Delivery Mode

```{r step2-restrict-samples}
tse_delivery_sp <- tse_delivery_sp[, !is.na(tse_delivery_sp$born_method)]

tse_delivery_sp$born_method <- factor(
  tse_delivery_sp$born_method,
  levels = c("vaginal", "c_section")
)
```

# Reduce to One Sample per Subject (Earliest Visit)

WampachL_2018 has repeated samples per infant (visits V1/V2/V3).
Keeping only the earliest sample per subject avoids pseudo-replication.

```{r step3-reduce-to-one-sample-per-subject}
meta_df <- as.data.frame(colData(tse_delivery_sp))
meta_df$sample_col <- colnames(tse_delivery_sp)

earliest_samples <- meta_df %>%
  group_by(subject_id) %>%
  slice_min(days_from_first_collection, n = 1, with_ties = FALSE) %>%
  pull(sample_col)

tse_delivery_sp <- tse_delivery_sp[, colnames(tse_delivery_sp) %in% earliest_samples]

# Sanity check - expect 16 samples, 16 subjects, 7 vaginal / 9 c_section
tse_delivery_sp
table(tse_delivery_sp$born_method)
n_distinct(colData(tse_delivery_sp)$subject_id)
```

# Alpha and Beta Diversity

```{r alpha-beta-diversity}
library(mia)
library(scater)
library(vegan)

# Alpha diversity (Shannon)
tse_delivery_sp <- addAlpha(
  tse_delivery_sp,
  assay.type = "relative_abundance",
  index = "shannon"
)

plotColData(
  tse_delivery_sp,
  y = "shannon",
  x = "born_method",
  colour_by = "born_method"
)

# Beta diversity (Bray-Curtis PCoA)
tse_delivery_sp <- runMDS(
  tse_delivery_sp,
  FUN = vegan::vegdist,
  method = "bray",
  assay.type = "relative_abundance",
  name = "PCoA_Bray"
)

plotReducedDim(
  tse_delivery_sp,
  dimred = "PCoA_Bray",
  color_by = "born_method"
)
```

# ANCOM-BC2

```{r method1-ancombc2}
out_delivery <- ancombc2(
  data = tse_delivery_sp,
  assay_name = "relative_abundance",
  fix_formula = "born_method"
)

res_delivery_ancombc2 <- out_delivery$res

diff_col <- grep("^diff_born_method", colnames(res_delivery_ancombc2), value = TRUE)
cat("ANCOM-BC2 significant taxa:", sum(res_delivery_ancombc2[[diff_col]], na.rm = TRUE), "\n")

knitr::kable(res_delivery_ancombc2)
```

# LEfSe

lefser needs its OWN relativeAb() (scales to sum to 1,000,000),
not a general-purpose transform - see project notes.

```{r method2-lefse}
delivery_ra <- lefser:::relativeAb(tse_delivery_sp)
delivery_ra$born_method <- droplevels(as.factor(delivery_ra$born_method))

res_delivery_lefse <- lefser(
  delivery_ra,
  classCol = "born_method",
  lda.threshold = 2,
  trim.names = TRUE
)

lefserPlot(res_delivery_lefse)
```

#  MAaSLin2

```{r method3-maaslin2}
maaslin_delivery_input <- as.data.frame(t(assay(tse_delivery_sp, "relative_abundance")))
maaslin_delivery_meta  <- as.data.frame(colData(tse_delivery_sp))

fit_delivery_maaslin <- Maaslin2(
  input_data      = maaslin_delivery_input,
  input_metadata  = maaslin_delivery_meta,
  output          = "maaslin_delivery_output",
  fixed_effects   = "born_method",
  normalization   = "TSS",
  min_prevalence  = 0.1
)

#Significant results: maaslin_delivery_output/significant_results.tsv

```

# Method 4: Additional

## Corncob

```{r method4-corncob}
ps_delivery <- mia::convertToPhyloseq(tse_delivery_sp, assay.type = "relative_abundance")

library(corncob)

corncob_delivery <- differentialTest(
  formula          = ~ born_method,
  phi.formula      = ~ born_method,
  formula_null     = ~ 1,
  phi.formula_null = ~ born_method,
  test = "Wald",
  data = ps_delivery,
  fdr_cutoff = 0.05
)

corncob_delivery$significant_taxa
```

## DESeq2

```{r method4-deseq2}
library(DESeq2)

counts_delivery <- round(assay(tse_delivery_sp, "relative_abundance"))

dds_delivery <- DESeqDataSetFromMatrix(
  countData = counts_delivery + 1L,   # pseudocount avoids all-zero rows
  colData   = as.data.frame(colData(tse_delivery_sp)),
  design    = ~ born_method
)

dds_delivery <- DESeq(dds_delivery)
res_deseq2_delivery <- results(dds_delivery)
head(res_deseq2_delivery[order(res_deseq2_delivery$padj), ])
```
