---
title: "SEVERITY ANALYSIS — BrooksB_2017"
subtitle: "Gestational age (continuous) among preterm infants"
author: "Sally Karimi Kinyua"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

<!--
##########################################################
## SEVERITY ANALYSIS — BrooksB_2017
## Gestational age (continuous) among preterm infants
## Methods: ANCOM-BC2, MAaSLin2 (+ optional DESeq2)
## NOTE: LEfSe is NOT applicable here - it requires exactly 2 groups,
## and gestational_age is used here as a continuous predictor.
##########################################################
-->

> **Methods:** ANCOM-BC2, MAaSLin2 (+ optional Corncob, DESeq2)
>
> **Note:** LEfSe is NOT applicable here — it requires exactly 2 groups, and `gestational_age` is used here as a continuous predictor.

# Setup

```{r setup}
library(curatedMetagenomicData)
library(dplyr)
library(ANCOMBC)
library(Maaslin2)
```

# Download Data

```{r step1-download-data}
brooks_data <- curatedMetagenomicData(
  "BrooksB_2017.relative_abundance",
  dryrun = FALSE,
  counts = TRUE   # counts scale needed for ANCOM-BC2/MAaSLin2/DESeq2
)

tse_severity_sp <- brooks_data[[1]]
```

# Restrict to Preterm Infants with Known Gestational Age

```{r step2-restrict-samples}
tse_severity_sp <- tse_severity_sp[, tse_severity_sp$study_condition == "premature_born"]
tse_severity_sp <- tse_severity_sp[, !is.na(tse_severity_sp$gestational_age)]
```

# Check Alpha and Beta Diversity

```{r alpha-beta-diversity}
library(mia)
library(scater)
library(vegan)

# Alpha diversity (Shannon)
tse_severity_sp <- addAlpha(
  tse_severity_sp,
  assay.type = "relative_abundance",
  index = "shannon"
)

# Continuous predictor (gestational_age) - use a scatter, not a boxplot
plotColData(
  tse_severity_sp,
  y = "shannon",
  x = "gestational_age"
) + ggplot2::geom_smooth(method = "lm", se = TRUE)

# Beta diversity (Bray-Curtis PCoA)
tse_severity_sp <- runMDS(
  tse_severity_sp,
  FUN = vegan::vegdist,
  method = "bray",
  assay.type = "relative_abundance",
  name = "PCoA_Bray"
)

plotReducedDim(
  tse_severity_sp,
  dimred = "PCoA_Bray",
  colour_by = "gestational_age"
)
```

# Reduce to One Sample per Subject (Earliest Visit)

BrooksB_2017 has 403 samples from only 29 subjects (longitudinal).
Keeping only the earliest sample per subject avoids pseudo-replication.

```{r step3-reduce-to-one-sample-per-subject}
meta_sev <- as.data.frame(colData(tse_severity_sp))
meta_sev$sample_col <- colnames(tse_severity_sp)

earliest_sev <- meta_sev %>%
  group_by(subject_id) %>%
  slice_min(days_from_first_collection, n = 1, with_ties = FALSE) %>%
  pull(sample_col)

tse_severity_sp <- tse_severity_sp[, colnames(tse_severity_sp) %in% earliest_sev]

# Sanity check - expect ~29 samples, one per subject
tse_severity_sp
summary(tse_severity_sp$gestational_age)
n_distinct(colData(tse_severity_sp)$subject_id)
```

# ANCOM-BC2 (Continuous Predictor)

```{r method1-ancombc2}
out_severity <- ancombc2(
  data = tse_severity_sp,
  assay_name = "relative_abundance",
  fix_formula = "gestational_age"
)

res_severity_ancombc2 <- out_severity$res

diff_col <- grep("^diff_gestational_age", colnames(res_severity_ancombc2), value = TRUE)
cat("ANCOM-BC2 significant taxa:", sum(res_severity_ancombc2[[diff_col]], na.rm = TRUE), "\n")

knitr::kable(res_severity_ancombc2)
```

#  MAaSLin2 (Continuous Predictor)

```{r method2-maaslin2}
maaslin_severity_input <- as.data.frame(t(assay(tse_severity_sp, "relative_abundance")))
maaslin_severity_meta  <- as.data.frame(colData(tse_severity_sp))

fit_severity_maaslin <- Maaslin2(
  input_data      = maaslin_severity_input,
  input_metadata  = maaslin_severity_meta,
  output          = "maaslin_severity_output",
  fixed_effects   = "gestational_age",
  normalization   = "TSS",
  min_prevalence  = 0.1
)
```

# Additional Methods

## Corncob

```{r method3-corncob}
ps_severity <- mia::convertToPhyloseq(tse_severity_sp, assay.type = "relative_abundance")

library(corncob)

corncob_severity <- differentialTest(
  formula          = ~ gestational_age,
  phi.formula      = ~ gestational_age,
  formula_null     = ~ 1,
  phi.formula_null = ~ gestational_age,
  test = "Wald",
  data = ps_severity,
  fdr_cutoff = 0.05
)

corncob_severity$significant_taxa
```

## DESeq2

```{r method3-deseq2}
library(DESeq2)

counts_severity <- round(assay(tse_severity_sp, "relative_abundance"))

dds_severity <- DESeqDataSetFromMatrix(
  countData = counts_severity + 1L,
  colData   = as.data.frame(colData(tse_severity_sp)),
  design    = ~ gestational_age
)

dds_severity <- DESeq(dds_severity)
res_deseq2_severity <- results(dds_severity)
head(res_deseq2_severity[order(res_deseq2_severity$padj), ])

```