---
title: "Exploratory Analysis of Women's Health Microbial Signatures"
subtitle: "General Landscape Analysis (BugSigDB)"
author: "Sally Karimi Kinyua"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploratory Analysis of Women's Health Microbial Signatures}
  %\VignetteEngine{knitr::rmarkdown}
---

# Introduction

This vignette demonstrates an exploratory analysis of microbial signatures associated
with women's health conditions using data obtained from BugSigDB and functions provided
by the bugSigSimple package. This is the **General** analysis — a broad landscape view
across many women's-health-related conditions (breast cancer, cervical cancer,
endometriosis, pregnancy/maternal-child topics, and other reproductive/gynecological
conditions). Four focused sub-analyses (Cervical Cancer, Breast Cancer, Endometriosis,
and Pregnancy/Maternal-Child) build on this General analysis with condition-specific
rigor.

# Analysis objectives

1. Downloading the data from BugSigDB
2. Filtering the data to select women's-health-related signatures and remove duplicates
3. Description of the studies (body site, sequencing platform, publication year)
4. Description of the taxa
5. Core consensus signature and basic statistical tests
6. Monte Carlo simulation to calculate critical N
7. Heatmaps and dendrograms

# Environment note

**Before running this vignette**, confirm your Bioconductor infrastructure loads cleanly.
If `library(S4Vectors)` (or any of its dependents) fails with an error like *"package
could not be loaded"* or *"reached elapsed time limit"*, reinstall it fresh in an
otherwise-idle R session before proceeding:

```{r, eval=FALSE}
remove.packages("S4Vectors")
BiocManager::install("S4Vectors", update = FALSE, force = TRUE)
# Restart R (Session > Restart R) before continuing
```

# Making sure packages are installed

```{r, eval=FALSE, message=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("devtools", "tidyverse", "kableExtra"))
BiocManager::install(c("waldronlab/bugSigSimple", "waldronlab/BugSigDBStats", "waldronlab/bugsigdbr"))
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(bugSigSimple)
library(bugsigdbr)
library(BugSigDBStats)
library(tidyverse)
library(kableExtra)
library(ComplexHeatmap)
```

# Download BugSigDB data

```{r download-data}
full.dat <- bugsigdbr::importBugSigDB(cache = FALSE)

dim(full.dat)
```

# Remove empty signatures

```{r clean-data}
ind1 <- lengths(full.dat[["MetaPhlAn taxon names"]]) > 0
ind2 <- lengths(full.dat[["NCBI Taxonomy IDs"]]) > 0

dat <- full.dat[ind1 & ind2, ]

dim(dat)
```

# Select women's-health-related signatures

```{r select-womens-health}
womens_health_terms <- c(
  # Breast
  "Breast cancer", "Breast cancer stage", "Breast carcinoma", "HER2 Positive Breast Carcinoma",
  # Cervical
  "Cervical cancer", "Cervical cerclage", "Cervical glandular intraepithelial neoplasia",
  "Cervical glandular intraepithelial neoplasia,Cervical cancer",
  "Cervical glandular intraepithelial neoplasia,Human papilloma virus infection",
  "Cervix erosion", "Human papilloma virus infection", "Human papilloma virus infection,Age",
  "Human papilloma virus infection,Cervical cancer",
  "Human papilloma virus infection,Cervical glandular intraepithelial neoplasia",
  "Chlamydophila infectious disease,Human papilloma virus infection",
  # Endometriosis
  "Endometriosis", "Adenomyosis", "Adenomyosis,Endometriosis",
  # Pregnancy / maternal-child
  "Pregnancy", "Gestational age", "Gestational diabetes", "Vaginal microbiome measurement",
  "Premature birth", "Spontaneous preterm birth", "Preterm premature rupture of the membranes",
  "Birth weight", "Birth measurement", "HIV mother to child transmission",
  "Neonatal death", "Neonatal jaundice", "Postnatal growth retardation",
  "Necrotizing enterocolitis", "Breastfeeding duration", "Breastfeeding duration,Transport",
  "Maternal milk", "Cesarean section", "Delivery method",
  # Broader women's health
  "Ovarian cancer", "Polycystic ovary syndrome", "Menopause", "Postmenopausal",
  "Postmenopausal osteoporosis", "Acquired primary ovarian failure", "Endometrial cancer",
  "Bacterial vaginosis", "Vaginitis", "Vulvovaginitis", "Vulvovaginal candidiasis",
  "Gonorrhea", "Chlamydia trachomatis", "Chlamydia trachomatis infectious disease",
  "Genital neoplasm, femaale", "Age at menarche", "Hormone replacement therapy", "Parous","Rosacea", "Ehlers-Danlos syndrome"
)


womens_health_dat <- dat %>%
  filter(Condition %in% womens_health_terms)

dim(womens_health_dat)

womens_health_dat %>%
  count(Condition, sort = TRUE) %>%
  knitr::kable()
```

Unlike the other Summer Projects analysis, no single body site captures the relevant evidence here —
vaginal, cervical, breast, and fecal samples are all legitimately relevant to different
women's-health conditions. No body-site restriction is applied in the General analysis;
body site distribution is instead described below as part of the study characteristics.

One study (conjunctival trachoma, caused by *Chlamydia trachomatis*) is retained as a
legitimate — if biologically atypical — result of the `Chlamydia trachomatis` filter term,
and is noted here as a minor caveat rather than excluded.

# Remove duplicate signatures

```{r remove-duplicates}
womens_health_dat <- womens_health_dat %>%
  distinct()

nrow(womens_health_dat)
```

No exact duplicate records were found in this signature set.

# Study characteristics

```{r study-summary}
study.summary <- womens_health_dat %>%
  count(`Body site`, `Sequencing type`)

knitr::kable(study.summary)

womens_health_dat %>%
  count(Year) %>%
  arrange(Year) %>%
  knitr::kable()
```

Body site distribution spans fecal samples (largely gestational diabetes, PCOS, and
breast cancer studies), reproductive-tract sites (vagina, cervix, endometrium - largely
bacterial vaginosis, HPV, and cervical cancer studies), and smaller categories including
meconium (mother-to-child transmission studies) and breast/milk tissue. Sequencing type
is heavily 16S-dominated (528 of 579 signatures), with a much smaller WMS subset (44) -
a real limitation to note when interpreting the WMS vs. 16S concordance check below, which
will be comparatively underpowered relative to the CRC analysis's more balanced platform
split. Publication years span 2010-2024, with the bulk of studies from 2018-2022; the
apparent taper in 2023-2024 most likely reflects BugSigDB curation lag rather than a
genuine decline in research activity.

# Taxonomic summary

```{r taxa-summary}
taxa.table <- createTaxonTable(womens_health_dat)

head(taxa.table)

knitr::kable(head(taxa.table, 20))
```

# Split by sequencing type

```{r split-sequencing}
wh.wms <- womens_health_dat %>% filter(`Sequencing type` == "WMS")
wh.16s <- womens_health_dat %>% filter(`Sequencing type` == "16S")

cat("WMS  - signatures:", nrow(wh.wms), "| studies:", n_distinct(wh.wms$PMID), "\n")
cat("16S  - signatures:", nrow(wh.16s), "| studies:", n_distinct(wh.16s$PMID), "\n")
```

# Long-format taxon table

```{r taxon-long}
sigs.womens_health <- bugsigdbr::getSignatures(womens_health_dat, tax.id.type = "taxname")

taxon.long <- purrr::map_dfr(seq_along(sigs.womens_health), function(i) {
  taxa <- sigs.womens_health[[i]]
  if (length(taxa) == 0) return(NULL)
  tibble::tibble(sig_id = i, taxon = taxa)
}) %>%
  left_join(
    womens_health_dat %>%
      mutate(sig_id = row_number()) %>%
      select(sig_id, PMID, Condition, `Sequencing type`, `Abundance in Group 1`),
    by = "sig_id"
  ) %>%
  rename(direction = `Abundance in Group 1`)

nrow(taxon.long)
```

# Taxon frequency across studies

```{r taxon-freq}
taxon.freq <- taxon.long %>%
  distinct(PMID, taxon, direction) %>%
  count(taxon, direction, name = "n_studies") %>%
  group_by(taxon) %>%
  mutate(total_studies = sum(n_studies)) %>%
  ungroup() %>%
  arrange(desc(total_studies))

n_distinct(taxon.freq$taxon)

taxon.freq %>%
  distinct(taxon, total_studies) %>%
  slice_max(total_studies, n = 15) %>%
  knitr::kable()
```

Given that General pools many biologically distinct conditions together (vaginal
dysbiosis, gut dysbiosis in metabolic conditions, breast tissue changes), the most
frequent taxa are expected to reflect a mix of vaginal-health organisms (e.g.
*Lactobacillus*, *Gardnerella*, *Prevotella*) and gut-health organisms (e.g.
*Bacteroides*, *Faecalibacterium*) rather than a single coherent signature - this
heterogeneity is itself part of the motivation for the four focused sub-analyses.

# Core consensus signature

```{r core-signature}
concordance <- taxon.long %>%
  distinct(PMID, taxon, direction) %>%
  group_by(taxon) %>%
  summarise(
    n_studies       = n_distinct(PMID),
    n_increased     = sum(direction == "increased"),
    n_decreased     = sum(direction == "decreased"),
    concordance_pct = round(max(n_increased, n_decreased) / n_studies * 100, 1),
    majority_dir    = if_else(n_increased >= n_decreased, "increased", "decreased"),
    .groups = "drop"
  ) %>%
  arrange(desc(n_studies), desc(concordance_pct))

core.sig <- concordance %>%
  filter(n_studies >= 3, concordance_pct >= 75) %>%
  arrange(majority_dir, desc(n_studies))

nrow(core.sig)

knitr::kable(core.sig)
```

```{r core-signature-plot, fig.height=max(9, 0.45 * 15 + 3), fig.width=9}
core.sig %>%
  group_by(majority_dir) %>%
  slice_max(n_studies, n = 15, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    direction_label = if_else(majority_dir == "increased",
                               "Increased", "Decreased"),
    taxon_label = paste0(str_wrap(taxon, width = 28), " (", n_studies, ")")
  ) %>%
  ggplot(aes(x = reorder(taxon_label, n_studies), y = n_studies, fill = direction_label)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(concordance_pct, "%")), hjust = -0.15, size = 3.3) +
  coord_flip(clip = "off") +
  facet_wrap(~direction_label, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c("Increased" = "#800020", "Decreased" = "#FFCBA4")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.35))) +
  labs(
    title = "Core Women's Health Consensus Signature (General)",
    subtitle = "Top 15 taxa per direction, in >=3 studies with >=75% direction concordance",
    x = NULL, y = "Number of independent studies", fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 10, lineheight = 0.9),
    strip.text = element_text(size = 11, face = "bold"),
    plot.margin = margin(t = 10, r = 20, b = 10, l = 10)
  )
```


```{r core-signature-full-table}
knitr::kable(core.sig)
```

```{r body-site-plot, fig.height=7, fig.width=9}
womens_health_dat %>%
  count(`Body site`, sort = TRUE) %>%
  filter(n >= 2) %>%
  ggplot(aes(x = reorder(`Body site`, n), y = n)) +
  geom_col(fill = "#C154C1") +
  coord_flip() +
  labs(
    title = "Women's Health Signatures by Body Site",
    subtitle = "Sites with at least 2 signatures",
    x = NULL, y = "Number of signatures"
  ) +
  theme_minimal(base_size = 11)
```



# WMS vs 16S concordance

```{r wms-16s-concordance}
dir.by.platform <- taxon.long %>%
  filter(`Sequencing type` %in% c("WMS", "16S")) %>%
  distinct(PMID, taxon, direction, `Sequencing type`) %>%
  count(taxon, `Sequencing type`, direction, name = "n_studies") %>%
  group_by(taxon, `Sequencing type`) %>%
  slice_max(n_studies, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(taxon, `Sequencing type`, majority_dir = direction)

platform.compare <- dir.by.platform %>%
  pivot_wider(names_from = `Sequencing type`, values_from = majority_dir) %>%
  filter(!is.na(WMS), !is.na(`16S`)) %>%
  rename(dir_WMS = WMS, dir_16S = `16S`) %>%
  mutate(agree = dir_WMS == dir_16S)

nrow(platform.compare)
sum(platform.compare$agree)
round(100 * mean(platform.compare$agree), 1)

knitr::kable(platform.compare)
```

Note the much smaller WMS sample here (44 signatures total, vs. hundreds of 16S) compared
to the CRC analysis - this comparison is real but less statistically powered than its CRC
counterpart, and should be presented with that caveat.

# Taxon enrichment analysis

```{r taxon-enrichment}
n.universe <- n_distinct(taxon.long$taxon)
n.sigs     <- length(sigs.womens_health)

enrichment.df <- taxon.freq %>%
  distinct(taxon, total_studies) %>%
  mutate(
    observed         = total_studies,
    expected         = n.sigs / n.universe,
    enrichment_ratio = observed / expected,
    log2_enrichment  = log2(enrichment_ratio),
    p_value = purrr::map_dbl(observed, ~ binom.test(
      x = .x, n = n.sigs, p = 1 / n.universe, alternative = "greater"
    )$p.value),
    p_adj = p.adjust(p_value, method = "BH")
  ) %>%
  arrange(p_adj, desc(log2_enrichment))

sum(enrichment.df$p_adj < 0.05)

knitr::kable(head(enrichment.df, 15))
```
```{r geographic-trends, fig.height=8, fig.width=9}
womens_health_dat %>%
  distinct(PMID, `Location of subjects`, `Sequencing type`) %>%
  count(`Location of subjects`, `Sequencing type`) %>%
  ggplot(aes(x = reorder(`Location of subjects`, n), y = n, fill = `Sequencing type`)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Women's Health Studies by Reported Location",
    x = NULL, y = "Number of studies"
  ) +
  theme_minimal(base_size = 11)
```

```{r temporal-trends, fig.height=6, fig.width=8}
womens_health_dat %>%
  distinct(PMID, Year, `Sequencing type`) %>%
  count(Year, `Sequencing type`) %>%
  filter(!is.na(Year)) %>%
  group_by(`Sequencing type`) %>%
  arrange(Year) %>%
  mutate(cumulative_n = cumsum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = Year, y = cumulative_n, color = `Sequencing type`)) +
  geom_line(linewidth = 1.2) +
  geom_point() +
  labs(
    title = "Cumulative Women's Health Studies Over Time",
    x = "Publication year", y = "Cumulative number of studies", color = "Sequencing type"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```


# Monte Carlo critical N

```{r monte-carlo}
siglengths <- lengths(womens_health_dat$`MetaPhlAn taxon names`)

criticalN <- getCriticalN(
  relevant.sigs = womens_health_dat$`MetaPhlAn taxon names`,
  siglengths = siglengths
)

criticalN
```

```{r criticalN-plot, fig.height=5, fig.width=6}
hist(
  criticalN$simulated_values,
  breaks = 30,
  main = "Monte Carlo Simulation of Critical N (Women's Health, General)",
  xlab = "Maximum frequency under the null model"
)

abline(
  v = criticalN$critical_n,
  col = "#FF00FF",
  lwd = 2,
  lty = 2
)
```

# Signature similarity

```{r signature-similarity}
allsigs <- bugsigdbr::getSignatures(womens_health_dat, tax.id.type = "taxname")

allsigs <- allsigs[sapply(allsigs, length) > 1]

length(allsigs)

mydists <- BugSigDBStats::calcPairwiseOverlaps(allsigs)

dim(mydists)

jmat <- BugSigDBStats::calcJaccardSimilarity(allsigs)

row_labels <- sub("^bsdb:", "PMID", sub("_.*$", "", rownames(jmat)))
col_labels <- sub("^bsdb:", "PMID", sub("_.*$", "", colnames(jmat)))
```

# Heatmap



```{r heatmap, fig.height=7, fig.width=9}
library(grid)
library(ComplexHeatmap)

siglengths <- sapply(allsigs, length)

# Set standard, crisp base font sizes
ha <- HeatmapAnnotation(
  `Signature\nLength` = anno_barplot(siglengths, gp = gpar(fill = "steelblue"))
)

hr <- rowAnnotation(
  `Signature\nLength` = anno_barplot(siglengths, gp = gpar(fill = "steelblue"))
)

hm <- Heatmap(
  jmat,
  top_annotation = ha,
  left_annotation = hr,
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_title = "Heatmap of Similarity of Women's Health Microbial Signatures",
  column_title_gp = gpar(fontsize = 12, fontface = "bold"),
  heatmap_legend_param = list(title = "Jaccard Similarity", direction = "horizontal")
)

# Draw cleanly directly to your HTML page output
draw(hm, heatmap_legend_side = "bottom")

# Force turn off any weird existing devices
dev.off() 

# Open clean canvas with safe, static dimensions
png(
  filename = "WomensHealth_heatmap.png",
  width = 12,
  height = 10,
  units = "in",
  res = 300
)

# Draw the heatmap inside the device
draw(
  hm,
  heatmap_legend_side = "bottom",
  annotation_legend_side = "right",
  padding = unit(c(2, 2, 2, 10), "mm")
)

# Crucial: turn off the device to save the file cleanly
dev.off()

```

Given the heterogeneity already observed in the taxon frequency table (vaginal-health taxa
vs. gut-health taxa), this heatmap is expected to reveal distinct clusters of studies
rather than one homogeneous signature - likely separating along body-site/condition lines
(vaginal/cervical studies vs. fecal/metabolic studies) rather than a single unified
women's-health signature. This heterogeneity is the core rationale for the four focused
sub-analyses that follow this General analysis.

# Dendrogram

```{r dendrogram, fig.height=6, fig.width=9}
hc <- hclust(as.dist(1 - jmat))
dend <- as.dendrogram(hc)

# Standard margin structure
par(mar = c(4, 1, 3, 1))

plot(
  dend,
  horiz = TRUE,
  leaflab = "none",
  main = "Hierarchical Cluster Map of Women's Health Microbial Signatures",
  xlab = "Distance (1 - Jaccard similarity)",
  cex.main = 1.1
)
```
11}


```

# Discussion

This exploratory analysis identified `r nrow(womens_health_dat)` women's-health-related
microbial signatures curated in BugSigDB, spanning breast cancer, cervical cancer,
endometriosis, pregnancy/maternal-child topics, and broader gynecological conditions
(PCOS, menopause, bacterial vaginosis, and related infections).

Key observations from this General analysis:

- A **formal core-signature threshold** (>=3 independent studies, >=75% directional
  concordance) identified `r nrow(core.sig)` taxa with consistent evidence across the
  whole women's-health evidence base.
- The most frequently reported taxa split cleanly into two biologically distinct clusters:
  **vaginal-health organisms** (*Lactobacillus*, *Gardnerella*, *Prevotella*, *Sneathia*,
  *Megasphaera*, *Dialister*) associated largely with bacterial vaginosis, HPV, and
  cervical cancer studies, and **gut-health organisms** (*Bacteroides*,
  *Faecalibacterium*, *Blautia*) associated largely with gestational diabetes, PCOS, and
  breast cancer studies.
- **WMS vs. 16S concordance** is reported, but is comparatively underpowered relative to
  the CRC analysis given the strong 16S skew (528 vs. 44 signatures) in this dataset.
- **Body site distribution** confirms that, unlike CRC, no single body site captures the
  relevant evidence - vaginal, cervical, fecal, and breast tissue samples are all
  legitimately part of the women's-health evidence base.
- One study (conjunctival trachoma) was retained as a legitimate but biologically atypical
  result of the `Chlamydia trachomatis` filter term - a minor caveat rather than an error.
- BugSigDB currently has **no dedicated postpartum category** and only one study with a
  clean postpartum-specific comparison (maternal vs. infant samples at 6 weeks
  postpartum) - too sparse for formal statistical treatment, and mentioned here only as a
  gap in the current evidence base.

The biological heterogeneity uncovered here - two largely distinct microbial communities
(vaginal vs. gut) responding to different conditions - is the central motivation for the
four focused sub-analyses (Cervical Cancer, Breast Cancer, Endometriosis, and
Pregnancy/Maternal-Child) that follow, each of which applies this same rigor within a
single, biologically coherent condition.
