---
title: "Focused Sub-Analysis of Pregnancy & Maternal-Child Microbial Signatures"
subtitle: "Vaginal/Gut Changes in Pregnancy, Preterm Birth, and Mother-to-Child Transmission (BugSigDB)"
author: "Sally Karimi Kinyua"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Focused Sub-Analysis of Pregnancy and Maternal-Child Microbial Signatures}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(bugSigSimple)
library(bugsigdbr)
library(BugSigDBStats)
library(tidyverse)
library(kableExtra)
library(ComplexHeatmap)
```

# Introduction

This vignette focuses on pregnancy and maternal-child microbiome signatures curated in
BugSigDB, spanning three internal groupings: (1) vaginal/gut microbiome changes during
pregnancy itself, (2) preterm birth-associated signatures, and (3) neonatal/infant gut
microbiome signatures reflecting mother-to-child transmission. A fourth topic,
**postpartum** microbiome changes, is deliberately **not** included as a formal
sub-grouping here: BugSigDB currently has no dedicated postpartum condition term, and a
keyword search across study titles/keywords/group names turned up only a single clean
postpartum-specific comparison (Study 617: maternal vs. infant samples at 6 weeks
postpartum) - too sparse for formal statistical treatment. It is mentioned only as a
single case-study example in the Discussion section below.

# Download and Prepare Data

```{r download-data}
full.dat <- bugsigdbr::importBugSigDB(cache = FALSE)

ind1 <- lengths(full.dat[["MetaPhlAn taxon names"]]) > 0
ind2 <- lengths(full.dat[["NCBI Taxonomy IDs"]]) > 0
dat <- full.dat[ind1 & ind2, ]
```

# Select Pregnancy/Maternal-Child Signatures

```{r select-pregnancy}
pregnancy_vaginal_gut_terms <- c(
  "Pregnancy", "Gestational age", "Gestational diabetes", "Vaginal microbiome measurement"
)

preterm_birth_terms <- c(
  "Premature birth", "Spontaneous preterm birth", "Preterm premature rupture of the membranes",
  "Birth weight", "Birth measurement"
)

neonatal_mother_child_terms <- c(
  "HIV mother to child transmission", "Neonatal death", "Neonatal jaundice",
  "Postnatal growth retardation", "Necrotizing enterocolitis", "Breastfeeding duration",
  "Breastfeeding duration,Transport", "Maternal milk", "Cesarean section", "Delivery method"
)

pregnancy_terms <- c(
  pregnancy_vaginal_gut_terms, preterm_birth_terms, neonatal_mother_child_terms
)

pregnancy_dat <- dat %>%
  filter(Condition %in% pregnancy_terms) %>%
  distinct()

nrow(pregnancy_dat)

pregnancy_dat %>%
  count(Condition, sort = TRUE) %>%
  knitr::kable()
```

## Breakdown by Internal Grouping

```{r subgroup-breakdown}
pregnancy_dat <- pregnancy_dat %>%
  mutate(
    subgroup = case_when(
      Condition %in% pregnancy_vaginal_gut_terms   ~ "Vaginal/Gut During Pregnancy",
      Condition %in% preterm_birth_terms           ~ "Preterm Birth",
      Condition %in% neonatal_mother_child_terms   ~ "Neonatal/Mother-to-Child",
      TRUE ~ "Other"
    )
  )

pregnancy_dat %>%
  count(subgroup, sort = TRUE) %>%
  knitr::kable(caption = "Signatures per Internal Grouping")
```

# Study Characteristics

```{r study-summary}
study.summary <- pregnancy_dat %>%
  count(`Body site`, `Sequencing type`)

knitr::kable(study.summary)

pregnancy_dat %>%
  count(Year) %>%
  arrange(Year) %>%
  knitr::kable()
```

```{r body-site-plot, fig.height=6, fig.width=8}
pregnancy_dat %>%
  count(`Body site`, sort = TRUE) %>%
  filter(n >= 2) %>%
  ggplot(aes(x = reorder(`Body site`, n), y = n)) +
  geom_col(fill = "#FFD1DC", width = 0.6) +
  coord_flip() +
  labs(
    title = "Pregnancy/Maternal-Child Signatures by Body Site",
    subtitle = "Sites with at least 2 signatures",
    x = NULL, y = "Number of Signatures"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

```{r subgroup-plot, fig.height=5, fig.width=8}
pregnancy_dat %>%
  count(subgroup, `Sequencing type`) %>%
  ggplot(aes(x = reorder(subgroup, n), y = n, fill = `Sequencing type`)) +
  geom_col() +
  coord_flip() +
  # Provided 3 distinct pastel colors to match your 3 sequencing categories
  scale_fill_manual(values = c("#FFB6C1", "#89CFF0", "#A3E4D7")) + 
  labs(
    title = "Signatures by Internal Grouping and Sequencing Platform",
    x = NULL, y = "Number of Signatures", fill = "Platform"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

```{r geographic-trends, fig.height=7, fig.width=8}
pregnancy_dat %>% 
  distinct(PMID, `Location of subjects`, `Sequencing type`) %>% 
  filter(!is.na(`Location of subjects`)) %>% 
  count(`Location of subjects`, `Sequencing type`) %>% 
  ggplot(aes(x = reorder(`Location of subjects`, n), y = n, fill = `Sequencing type`)) + 
  geom_col() + 
  coord_flip() + 
  # Added emerald green to match the 3 required categories
  scale_fill_manual(values = c("#FBCCE7", "#A6C8FF", "#50C878")) +
  labs(
    title = "Pregnancy/Maternal-Child Studies by Reported Location", 
    x = NULL, 
    y = "Number of studies"
  ) + 
  theme_minimal(base_size = 11)
```

```{r temporal-trends, fig.height=6, fig.width=8}
pregnancy_dat %>%
  distinct(PMID, Year, `Sequencing type`) %>%
  filter(!is.na(Year)) %>%
  count(Year, `Sequencing type`) %>%
  group_by(`Sequencing type`) %>%
  arrange(Year) %>%
  mutate(cumulative_n = cumsum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = Year, y = cumulative_n, color = `Sequencing type`)) +
  geom_line(linewidth = 1.2) +
  geom_point() +
  labs(
    title = "Cumulative Pregnancy/Maternal-Child Studies Over Time",
    x = "Publication year", y = "Cumulative number of studies", color = "Sequencing type"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

# Taxonomic Summary

```{r taxa-summary}
taxa.table <- createTaxonTable(pregnancy_dat)
knitr::kable(head(taxa.table, 20))
```

# Split by Sequencing Type

```{r split-sequencing}
preg.wms <- pregnancy_dat %>% filter(`Sequencing type` == "WMS")
preg.16s <- pregnancy_dat %>% filter(`Sequencing type` == "16S")

cat("WMS  - signatures:", nrow(preg.wms), "| studies:", n_distinct(preg.wms$PMID), "\n")
cat("16S  - signatures:", nrow(preg.16s), "| studies:", n_distinct(preg.16s$PMID), "\n")
```

# Long-Format Taxon Table

```{r taxon-long}
sigs.pregnancy <- bugsigdbr::getSignatures(pregnancy_dat, tax.id.type = "taxname")
sigs.all <- bugsigdbr::getSignatures(dat, tax.id.type = "taxname")

taxon.long <- purrr::map_dfr(seq_along(sigs.all), function(i) {
  taxa <- sigs.all[[i]]
  if (length(taxa) == 0) return(NULL)
  tibble::tibble(sig_id = i, taxon = taxa)
}) %>%
  left_join(
    dat %>%
      mutate(sig_id = row_number()) %>%
      select(sig_id, PMID, Condition, `Sequencing type`, `Abundance in Group 1`),
    by = "sig_id"
  ) %>%
  rename(direction = `Abundance in Group 1`)

nrow(taxon.long)
```

# Taxon Frequency Across Studies

```{r taxon-freq}
taxon.freq <- taxon.long %>%
  filter(Condition %in% pregnancy_terms) %>%
  distinct(PMID, taxon, direction) %>%
  count(taxon, direction, name = "n_studies") %>%
  group_by(taxon) %>%
  mutate(total_studies = sum(n_studies)) %>%
  ungroup() %>%
  arrange(desc(total_studies))

n_distinct(taxon.freq$taxon)

taxon.freq %>%
  distinct(taxon, total_studies) %>%
  slice_max(total_studies, n = 15) %>%
  knitr::kable()
```

# Core Consensus Signature

```{r core-signature}
concordance <- taxon.long %>%
  filter(Condition %in% pregnancy_terms) %>%
  distinct(PMID, taxon, direction) %>%
  group_by(taxon) %>%
  summarise(
    n_studies       = n_distinct(PMID),
    n_increased     = sum(direction == "increased"),
    n_decreased     = sum(direction == "decreased"),
    concordance_pct = round(max(n_increased, n_decreased) / n_studies * 100, 1),
    majority_dir    = if_else(n_increased >= n_decreased, "increased", "decreased"),
    .groups = "drop"
  ) %>%
  arrange(desc(n_studies), desc(concordance_pct))

# Standard threshold; loosen to >=2/>=70% (as done for Cervical Cancer) if this
# returns very few/zero rows, and note that deviation in the Discussion section.
core.sig <- concordance %>%
  filter(n_studies >= 3, concordance_pct >= 75) %>%
  arrange(majority_dir, desc(n_studies))

nrow(core.sig)

knitr::kable(core.sig)
```

# Core Signature Visualisation

```{r core-signature-plot, fig.height=6, fig.width=9}
if (nrow(core.sig) > 0) {
  core.sig %>%
    group_by(majority_dir) %>%
    slice_max(n_studies, n = 15, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(
      direction_label = if_else(majority_dir == "increased", "Increased", "Decreased"),
      taxon_label = paste0(str_wrap(taxon, width = 28), " (", n_studies, ")")
    ) %>%
    ggplot(aes(x = reorder(taxon_label, n_studies), y = n_studies, fill = direction_label)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = paste0(concordance_pct, "%")), hjust = -0.15, size = 3.3) +
    coord_flip(clip = "off") +
    facet_wrap(~direction_label, scales = "free_y", ncol = 1) +
    scale_fill_manual(values = c("Increased" = "#FFC0CB", "Decreased" = "#B0E0E6")) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.35))) +
    labs(
      title = "Core Pregnancy/Maternal-Child Consensus Signature",
      x = NULL, y = "Number of independent studies", fill = NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none", plot.title = element_text(face = "bold"))
} else {
  cat("No taxa met the core-signature threshold at this sample size.")
}
```

# WMS vs 16S Concordance

```{r wms-16s-concordance}
dir.by.platform <- taxon.long %>%
  filter(Condition %in% pregnancy_terms) %>%
  filter(`Sequencing type` %in% c("WMS", "16S")) %>%
  distinct(PMID, taxon, direction, `Sequencing type`) %>%
  count(taxon, `Sequencing type`, direction, name = "n_studies") %>%
  group_by(taxon, `Sequencing type`) %>%
  slice_max(n_studies, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(taxon, `Sequencing type`, majority_dir = direction)

platform.compare <- dir.by.platform %>%
  pivot_wider(names_from = `Sequencing type`, values_from = majority_dir) %>%
  filter(!is.na(WMS), !is.na(`16S`)) %>%
  rename(dir_WMS = WMS, dir_16S = `16S`) %>%
  mutate(agree = dir_WMS == dir_16S)

nrow(platform.compare)
if (nrow(platform.compare) > 0) {
  sum(platform.compare$agree)
  round(100 * mean(platform.compare$agree), 1)
  knitr::kable(platform.compare)
} else {
  cat("No taxa reported by both WMS and 16S in this pregnancy dataset.")
}
```

# Taxon Enrichment Analysis

```{r taxon-enrichment}
n.universe <- n_distinct(taxon.long %>% filter(Condition %in% pregnancy_terms) %>% pull(taxon))
n.sigs     <- length(sigs.pregnancy)

enrichment.df <- taxon.freq %>%
  distinct(taxon, total_studies) %>%
  mutate(
    observed         = total_studies,
    expected         = n.sigs / n.universe,
    enrichment_ratio = observed / expected,
    log2_enrichment  = log2(enrichment_ratio),
    p_value = purrr::map_dbl(observed, ~ binom.test(
      x = .x, n = n.sigs, p = 1 / n.universe, alternative = "greater"
    )$p.value),
    p_adj = p.adjust(p_value, method = "BH")
  ) %>%
  arrange(p_adj, desc(log2_enrichment))

sum(enrichment.df$p_adj < 0.05)

knitr::kable(head(enrichment.df, 15))
```

# Monte Carlo Critical N

```{r monte-carlo}
siglengths <- lengths(pregnancy_dat$`MetaPhlAn taxon names`)

criticalN <- getCriticalN(
  relevant.sigs = pregnancy_dat$`MetaPhlAn taxon names`,
  siglengths = siglengths
)

criticalN
```

```{r criticalN-plot, fig.height=5, fig.width=6}
hist(
  criticalN$simulated_values,
  breaks = 30,
  main = "Monte Carlo Simulation of Critical N (Pregnancy/Maternal-Child)",
  xlab = "Maximum frequency under the null model"
)

abline(v = criticalN$critical_n, col = "violet", lwd = 2, lty = 2)
```

# Cross-Condition Taxa Overlap Analysis

```{r cross-condition-overlap}
pregnancy_taxa <- taxon.long %>%
  filter(Condition %in% pregnancy_terms) %>%
  distinct(taxon) %>%
  pull(taxon)

shared_taxa_counts <- taxon.long %>%
  filter(taxon %in% pregnancy_taxa) %>%
  group_by(taxon) %>%
  summarise(
    total_womens_health_studies = n_distinct(PMID),
    distinct_conditions_linked = n_distinct(Condition),
    top_other_conditions = paste(
      head(setdiff(unique(Condition), pregnancy_terms), 3),
      collapse = ", "
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(distinct_conditions_linked))

shared_taxa_counts %>%
  slice_max(distinct_conditions_linked, n = 20, with_ties = FALSE) %>%
  knitr::kable(
    caption = "Top Pregnancy-Associated Taxa and Their Links to Other Conditions",
    col.names = c("Taxon Name", "Total WH Studies", "Total Distinct Conditions", "Other Conditions it Often Appears In")
  )
```

# Signature Similarity and Heatmap

```{r heatmap, fig.height=10, fig.width=12}
allsigs <- bugsigdbr::getSignatures(pregnancy_dat, tax.id.type = "taxname")
allsigs <- allsigs[sapply(allsigs, length) > 1]

if (length(allsigs) > 1) {
  jmat <- BugSigDBStats::calcJaccardSimilarity(allsigs)
  siglengths <- sapply(allsigs, length)

  ha <- HeatmapAnnotation(
    `Signature\nLength` = anno_barplot(siglengths, gp = gpar(fill = "#1E90FF"), height = unit(1.2, "cm"))
  )
  hr <- rowAnnotation(
    `Signature\nLength` = anno_barplot(siglengths, gp = gpar(fill = "#1E90FF"), width = unit(1.2, "cm"))
  )

  hm <- Heatmap(
    jmat,
    top_annotation = ha,
    left_annotation = hr,
    show_row_names = FALSE,
    show_column_names = FALSE,
    row_km = 3,
    column_km = 3,
    row_gap = unit(1.5, "mm"),
    column_gap = unit(1.5, "mm"),
    row_dend_width = unit(2.5, "cm"),
    column_dend_height = unit(2.5, "cm"),
    column_title = "Heatmap of Pregnancy/Maternal-Child Microbial Signatures",
    column_title_gp = gpar(fontsize = 14, fontface = "bold"),
    heatmap_legend_param = list(title = "Jaccard Similarity", direction = "horizontal", legend_width = unit(6, "cm"))
  )

  draw(hm, heatmap_legend_side = "bottom", padding = unit(c(5, 5, 5, 5), "mm"))

  png(filename = "Pregnancy_heatmap.png", width = 12, height = 10, units = "in", res = 300)
  draw(hm, heatmap_legend_side = "bottom", padding = unit(c(5, 5, 5, 5), "mm"))
  dev.off()
}
```

# Dendrogram

```{r dendrogram, fig.height=14, fig.width=10}
if (exists("jmat") && nrow(jmat) > 1) {
  hc <- hclust(as.dist(1 - jmat))
  dend <- as.dendrogram(hc)

  par(mar = c(5, 2, 4, 2))

  plot(
    dend,
    horiz = TRUE,
    leaflab = "none",
    main = "Hierarchical Cluster Map of Pregnancy/Maternal-Child Microbial Signatures",
    xlab = "Distance (1 - Jaccard similarity)",
    cex.main = 1.3,
    lwd = 0.8
  )
}
```

